---
title: "34-0"
author: "Nate Bean"
date: 2020-12-08
categories: ["R", "NFL", "Rams"]
tags: ["NFL", "Rams", "Halftime"]
summary: "How unusual is the Rams perfect record when leading at halftime?"
authors: [Nate Bean]
twitterImg: "/post/rams-halftime-leads/featured.png"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE, tidy = "styler" )
```

The Rams put an end to some troubling streaks in a reassuring 38-28 win against the Cardinals on Sunday. They scored at least 30 points for the first time since Week 5, a drought that was partially due to the struggles of Jared Goff after a [hot start](https://natebean.rbind.io/post/goff-short-passing/) to the season. But after tallying 10 turnovers over the previous 4 weeks, he delivered a zero turnover performance.

But the Rams win also continued trends that have long been present in LA. After taking a 14-7 lead into halftime, the Rams victory brought them to a eye-popping 34-0 when leading at halftime since Sean McVay took over in 2017. It's becoming increasingly common to hear this stat cited by announcers or see it at the bottom of the ticker, and for good reason. It's a truly impressive streak. 

However, I wanted to get a bit more context for it. How much better is this record than that of the rest of league over the same period? I'll attempt to shed a little light on that question in this short post, once again using data from nflfastR. 

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(nflfastR)

mcvay <- (2017:2020)

half_scores <- map_df(mcvay, function(x) {
  readRDS(url(
    paste0("https://raw.githubusercontent.com/guga31bb/nflfastR-data/master/data/play_by_play_", x, ".rds")
  )) %>%
    mutate(year= x)}) 

half_data <- half_scores %>% #there are several games with na possession, does this happen after a score?
  group_by(game_id, game_half) %>%
  mutate(obs= row_number()) %>%
  filter(obs== max(obs)-1) %>%
  ungroup() %>%
  select(game_id, posteam, defteam, week, game_half, posteam_score_post, defteam_score_post) %>%
  mutate(leader= case_when(
    posteam_score_post > defteam_score_post ~ posteam,
    posteam_score_post < defteam_score_post ~ defteam,
    posteam_score_post == defteam_score_post ~ "tie"),
    lead= abs(posteam_score_post- defteam_score_post)) %>%
  group_by(game_id) %>%
  mutate(half_num= row_number()) %>%
  ungroup()
#combine these??
halftime_record <- half_data %>%
  group_by(game_id) %>%
  mutate(winner= leader[half_num==max(half_num)]) %>%
  ungroup() %>%
  filter(game_half== "Half1") %>% 
  mutate(half_win= ifelse(leader==winner, 1, 0)) %>% 
  filter(is.na(half_win)==FALSE) #figure out what is causing NAs and fix

#their record when losing or tied
rams <- halftime_record %>%
  filter(posteam== "LA"| defteam=="LA")

tied <- rams %>%
  filter(game_half=="Half1", leader=="tie") %>%
  mutate(wins= ifelse(winner=="LA", 1, 0))


losing <- rams %>%
  filter(game_half=="Half1", leader !="tie", leader != "LA") %>%
  mutate(wins= ifelse(winner=="LA", 1, 0),
         wins= sum(wins),
         losses= n()- mean(wins))

nfl_winning_perc <- round(sum(halftime_record$half_win)/nrow(halftime_record), digits=1)
```

The Rams have won 43 regular season and playoff games since 2017. While they're undefeated after leading at halftime, they've been 3-5 when tied at half and 6-16 while trailing. These stats confirm what any football fan would know- the team that leads at the half tends to win. Indeed, across the NFL over the same time period, `r nfl_winning_perc`% of games were won by the team that led at half. By winning nearly 40% of games during which they trailed at half, the Rams have actually outperformed the rest of the league.

The Rams perfect record when leading at half is better than average as well, but how much better is their winrate than other top NFL teams? The table below shows the records of the 10 teams with the highest winning percentages after leading at half. While the Rams are the only perfect team over when leading at the half,
the data shows that the best teams across the league don't have a significantly worse record when leading at the half. 
```{r, message=FALSE, warning=FALSE}
library(gt)
library(teamcolors)

team_logos <- teams_colors_logos %>% select(team_abbr, team_logo_espn)

halftime_record %>%
  filter(leader != "tie") %>%
  group_by(leader) %>%
  summarise(first_half_leads= n(),
            wins_when_leading= sum(half_win),
            lead= round(mean(lead), digits=1)) %>%
  mutate(win_perc = round(wins_when_leading/first_half_leads, digits=2)) %>%
  arrange(desc(win_perc)) %>%
  #format data for the table
  slice(1:10) %>%
  inner_join(team_logos, by= c("leader"= "team_abbr")) %>%
  select(leader,team_logo_espn, first_half_leads, wins_when_leading, lead, win_perc) %>%
  #now table
  gt() %>%
  text_transform(
    location = cells_body(vars(team_logo_espn)),
    fn = function(x){
      web_image(x)}) %>%
  tab_header(
    title = "NFL Winrates",
    subtitle = "When Leading at Halftime") %>%
  fmt_percent(
    columns= vars(win_perc),
    decimals = 0) %>%
  cols_label(
    team_logo_espn= "",
    leader= "",
    first_half_leads= "Leads",
    wins_when_leading= "Wins",
    lead= "Avg. Lead",
    win_perc= "Winrate") %>%
  tab_options(
    column_labels.border.top.width = px(3),
    column_labels.border.bottom.color = "black",
    table.border.top.width = px(3))
```
\
Once we get outside the top 10 teams winning percentage after leading at the half quickly drops below 80%. The Rams are noticeably better than this bottom 2/3 of teams, but they're only a bit better the other top teams. The data shows this stat isn't a unique attribute of Sean McVay's team, but a product of being a consistently good team and having a bit of luck as well. The fact that the Rams defense had only allowed two touchdowns in the second half prior to the Cardinals game helped too. Still, the Rams perfect 34-0 remains a fun statistic even if the difference between them and other teams wouldn't pass a statistical significance test. Here's hoping for many more!